---
title: 'Immune profiling analysis'
author: 'Christoph Hafemeister'
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document:
    toc: TRUE
    toc_depth: 2
    toc_float: TRUE
    code_folding: "hide"
params:
  config_file: "config.yaml"
---

```{r}
# load project-specific parameters
config <- yaml::read_yaml(params$config_file)

suppressPackageStartupMessages({
  library('Matrix')
  library('dplyr')
  library('Seurat')
  library('sctransform')
  library('reshape2')
  library('ggplot2')
  library('patchwork')
  library('ggrepel')
  source(file.path(config$project_root, 'R', 'utils.r'))
})

#future::plan(strategy = "multiprocess", workers = 1)
#options(future.globals.maxSize = 8 * 1024 ^ 3)
#options(future.fork.enable = TRUE)

knitr::opts_chunk$set(comment = NA, fig.width = 5, fig.height = 3.5, out.width = '60%',
                      warning = TRUE, error = TRUE, echo = TRUE, message = TRUE,
                      dpi = 100)
options(dplyr.summarise.inform = FALSE)
options(tibble.width = Inf)
options(DT.options = list(paging = TRUE, searching = FALSE, autoWidth = TRUE))

old <- theme_set(theme_bw(base_size = 11))
set.seed(8569205)
start_time <- proc.time()['elapsed']
```

# Load scRNA-seq results
```{r}
# load results of cell type annotation and join with clonotype summary
s_md <- readRDS(file = file.path(config$out_root, '03_integrated_NFATc1_annotated_meta_data.rds'))
s_md$barcode <- rownames(s_md)
```


# Immune profiling analysis

## Read data; first overview

Read clonotype contig annotations into one big data frame and print table of Sample ID vs Chain detected. Note that we keep only observations with productive contig (likely functional T or B cell receptor) and annotated with a single chain (discarding 'Multi' chains that we see in gdTCR results quite often)
```{r}
path = config$data_root
clonotype_files <- list.files(path = path, pattern = '^filtered_contig_annotations\\.csv$', recursive = TRUE, full.names = TRUE)
ret <- list()
for (f in clonotype_files) {
  clono_dat <- read.csv(f)
  clono_dat$dir_name <- basename(dirname(f))
  # also read the clonotype file, so we can add the iNKT and MAIT evidence
  clono_anno <- read.csv(file = gsub(pattern = 'filtered_contig_annotations\\.csv$', 
                                     x = f,
                                     replacement = 'clonotypes.csv'))
  if (all(c('clonotype_id', 'inkt_evidence', 'mait_evidence') %in% colnames(clono_anno))) {
    clono_dat <- left_join(clono_dat, clono_anno[, c('clonotype_id', 'inkt_evidence', 'mait_evidence')], by = c('raw_clonotype_id' = 'clonotype_id'))
  } else {
    clono_dat$inkt_evidence <- ''
    clono_dat$mait_evidence <- ''
  }
  
  ret[[basename(dirname(f))]] <- clono_dat
}
clono_dat <- do.call(rbind, ret)
clono_dat <- do.call(rbind, ret) %>% filter(productive %in% c('true', 'True'), chain != 'Multi')
```

Table of directory name and chain
```{r}
print(table(clono_dat$dir_name, clono_dat$chain))
```

Infer sample IDs from directory names and show sample-chain table
```{r}
# infer sample ids from directory names
clono_dat$sample_id <- gsub(pattern = '_(abTCR|BCR|gdTCR)_VDJ(_GD)*', replacement = '', x = clono_dat$dir_name)
print(table(clono_dat$sample_id, clono_dat$chain))

clono_dat$sample_number <- as.integer(stringr::str_extract(clono_dat$sample_id, '\\d+'))
# make sure the cell barcode matches the names assigned by Seurat in the merged and integrated analysis
clono_dat$barcode <- paste(clono_dat$barcode, clono_dat$sample_number, sep = '_')
```
Keep only the relevant samples
```{r}
clono_dat <- clono_dat[clono_dat$sample_id %in% s_md$orig.ident, ]
```


Aggregate the chain counts per cell. How often do we see certain chains together (and how many of them)?
```{r}
chain_counts <- arrange(clono_dat, sample_id, barcode, chain) %>%
  group_by(sample_id, barcode, chain) %>% 
  summarise(n = n()) %>%
  mutate(chain_count = paste(chain, n, sep=':')) %>%
  summarise(chains = paste(chain_count, collapse = ';'))
tab <- as.data.frame(sort(table(chain_counts$chains), decreasing = TRUE))
colnames(tab) <- c('Chain counts', 'Cells')
DT::datatable(tab)
```


```{r}
# combine the information per cell; use only data with productive CDR3
clono_sum <- arrange(clono_dat, sample_id, barcode, chain) %>%
  group_by(sample_id, barcode) %>% 
  mutate(chain_cdr3 = paste(chain, cdr3, sep=':')) %>%
  summarise(clonotype = paste(chain_cdr3, collapse = ';'),
            inkt_evidence = paste(setdiff(sort(unique(inkt_evidence)), ''), collapse = ';'),
            mait_evidence = paste(setdiff(sort(unique(mait_evidence)), ''), collapse = ';')) %>%
  mutate(has_TRAB = grepl(pattern = 'TR[AB]:', x = clonotype),
         has_IG = grepl(pattern = 'IG[HKL]:', x = clonotype),
         has_TRDG = grepl(pattern = 'TR[DG]:', x = clonotype),
         only_TRAB = has_TRAB & !has_IG & !has_TRDG,
         only_IG = !has_TRAB & has_IG & !has_TRDG,
         only_TRDG = !has_TRAB & !has_IG & has_TRDG)

freq_ab <- filter(clono_sum, has_TRAB) %>%
  group_by(sample_id, clonotype) %>%
  summarise(n = n()) %>%
  mutate(freq = n / sum(n), rank = rank(-freq, ties.method = 'min')) %>% 
  arrange(sample_id, -freq)

freq_ig <- filter(clono_sum, has_IG) %>%
  group_by(sample_id, clonotype) %>%
  summarise(n = n()) %>%
  mutate(freq = n / sum(n), rank = rank(-freq, ties.method = 'min')) %>% 
  arrange(sample_id, -freq)

freq_dg <- filter(clono_sum, has_TRDG) %>%
  group_by(sample_id, clonotype) %>%
  summarise(n = n()) %>%
  mutate(freq = n / sum(n), rank = rank(-freq, ties.method = 'min')) %>% 
  arrange(sample_id, -freq)

# join with scRNA-seq results
s_md <- left_join(s_md, clono_sum, by = 'barcode')
pred_cell_types <- sort(unique(as.character(s_md$az.annotation)))
s_md$az.annotation <- factor(s_md$az.annotation, levels = pred_cell_types)
s_md <- left_join(s_md, chain_counts)
```

Save for later
```{r}
saveRDS(object = list(clono_dat = clono_dat, 
                      clono_sum = clono_sum,
                      md = s_md), 
        file = file.path(config$out_root, '04_clonotypes_NFATc1.rds'))
```


## TRA and TRB 

```{r, fig.width=10, fig.height=4.5, out.width='100%'}
p1 <- filter(s_md, has_TRAB) %>%
  ggplot(aes(sample_id, fill = chains)) + geom_bar() + ylab('Cells') +
  ggtitle('Cells with at least one A or B chain') +
  scale_fill_discrete(name = 'Chain counts') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
show(p1)
```

Clonotype diversity based on cells with exactly one A and one B chains observed
```{r}
tmp <- filter(s_md, chains == 'TRA:1;TRB:1') %>%
  group_by(sample_id) %>%
  summarise(n = n(), sdi(clonotype))
p2 <- ggplot(tmp, aes(n, exp(sdi))) + 
  geom_abline(slope = 1, color = 'gray80') +
  geom_point() +
  geom_label_repel(aes(label = sample_id), max.overlaps = Inf) +
  xlab('Cells') + ylab('Effective number of clonotypes') 
show(p2)

tmp <- filter(s_md, chains == 'TRA:1;TRB:1') %>%
  group_by(sample_id) %>%
  summarise(sdi_ds = sdi_ds(clonotype))
p3 <- ggplot(tmp, aes(sample_id, sdi_ds)) + geom_boxplot() +
  ylab('Shannon diversity index') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p4 <- ggplot(tmp, aes(sample_id, exp(sdi_ds))) + geom_boxplot() +
  ylab('Effective number of clonotypes') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
show((p3 + p4) + plot_annotation(title = 'Downsampled to 100 cells, 1000 bootstraps'))
```


Show only clonotypes that have been ranked in the top 5 and have been observed more than two times
```{r, fig.width=10, out.width='100%'}
top_clono <- filter(freq_ab, rank <= 5, n > 2) %>% pull(clonotype) %>% unique()
filter(freq_ab, clonotype %in% top_clono) %>% 
  DT::datatable(rownames = FALSE) %>% 
  DT::formatRound(columns = 'freq', digits = 3)
```
```{r, out.width='100%', fig.width=10, fig.height=7}
p2 <- filter(freq_ab, clonotype %in% top_clono) %>% ungroup() %>%
  mutate(sample_id = factor(sample_id, levels = rev(unique(sample_id))),
         clonotype = factor(clonotype, levels = unique(clonotype))) %>%
  ggplot(aes(clonotype, sample_id, fill = freq)) +
  geom_raster() +
  scale_fill_continuous(name = 'Frequency') +
  theme(axis.text.x = element_text(angle = 66, hjust = 1)) +
  ggtitle('Clonotype frequency heatmap')
show(p2)
```

#### Clonality

As an additional clonotype diversity measure, we calculate clonality (one minus normalized entropy, aka one minus Peilou’s evenness index).

This is based on all cells with clonotype information regardless of whether they are in the RNA-seq based analysis.

Use only TCRB, ignore other chains, use only cells with exactly one TCRB chain.

```{r, fig.width=10, out.width='100%', fig.height=3.5}
df <- mutate(clono_sum, n_TRB = stringr::str_count(string = clonotype, pattern = 'TRB:')) %>%
  filter(n_TRB == 1) %>%
  mutate(TRB = stringr::str_extract(string = clonotype, pattern = 'TRB:[A-Z]+')) %>%
  group_by(sample_id) %>%
  summarise(cells = n(), clonotypes = length(unique(TRB)), clonality = 1 - norm_entropy(TRB))
p1 <- ggplot(df, aes(sample_id, clonality)) + geom_bar(stat = 'identity') + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab('Sample') + ylab('Clonality')
p2 <- gridExtra::tableGrob(df, rows = NULL)
show((p1 + p2) + plot_annotation(title = 'Clonality (TCRB) using all cells with one TCRB chain'))
```


Use both TCRA and TCRB chains to define clonotype, use only cells with exactly one TCRA chain and one TCRB chain, ignore other chains.

```{r, fig.width=10, out.width='100%', fig.height=3.5}
df <- mutate(clono_sum, 
             n_TRB = stringr::str_count(string = clonotype, pattern = 'TRB:'),
             n_TRA = stringr::str_count(string = clonotype, pattern = 'TRA:')) %>%
  filter(n_TRB == 1, n_TRA == 1) %>%
  mutate(AB = paste(stringr::str_extract(string = clonotype, pattern = 'TRA:[A-Z]+'),
                    stringr::str_extract(string = clonotype, pattern = 'TRB:[A-Z]+'))) %>%
  group_by(sample_id) %>%
  summarise(cells = n(), clonotypes = length(unique(AB)), clonality = 1 - norm_entropy(AB))
p1 <- ggplot(df, aes(sample_id, clonality)) + geom_bar(stat = 'identity') + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab('Sample') + ylab('Clonality')
p2 <- gridExtra::tableGrob(df, rows = NULL)
show((p1 + p2) + plot_annotation(title = 'Clonality (A+B) using all cells with exactly one TCRA and TCRB chain and RNA-seq'))
```

Clonality per integration cluster

Cells used: Cells that are in RNA-seq data and have exactly one TRB chain
Clonotype definition: AA sequence of TRB

```{r, fig.width=8, out.width='80%', fig.height=5}
df <- filter(s_md, grepl('TRB:1', chains)) %>%
  mutate(TRB = stringr::str_extract(string = clonotype, pattern = 'TRB:[A-Z]+')) %>%
  group_by(sample_id, integration_cluster) %>%
  summarise(cells = n(), clonotypes = length(unique(TRB)), clonality = 1 - norm_entropy(TRB))

filter(df, cells > 9) %>%
  ggplot(aes(sample_id, clonality, size = cells)) + geom_point() +
  facet_wrap(~ integration_cluster, ncol = 5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position="bottom") +
  xlab('Sample ID') + ylab('Clonality') +
  ggtitle('Clonality per cluster (10 or more cells) per sample (TCRB only)')
```

Clonality per cell type (unfiltered Azimuth predictions)

Cells used: Cells that are in RNA-seq data and have exactly one TRB chain
Clonotype definition: AA sequence of TRB

```{r, fig.width=8, out.width='80%', fig.height=5}
df <- filter(s_md, grepl('TRB:1', chains)) %>%
  dplyr::rename(celltype = az.annotation) %>%
  mutate(TRB = stringr::str_extract(string = clonotype, pattern = 'TRB:[A-Z]+')) %>%
  group_by(sample_id, celltype) %>%
  summarise(cells = n(), clonotypes = length(unique(TRB)), clonality = 1 - norm_entropy(TRB))

filter(df, cells > 9) %>%
  ggplot(aes(sample_id, clonality, size = cells)) + geom_point() +
  facet_wrap(~ celltype, ncol = 5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position="bottom") +
  xlab('Sample ID') + ylab('Clonality') +
  ggtitle('Clonality per cell type (10 or more cells) per sample (TCRB only)')
```

Clonality per cell type (filtered Azimuth predictions)

Cells used: Cells that are in RNA-seq data and have exactly one TRB chain
Clonotype definition: AA sequence of TRB

```{r, fig.width=8, out.width='80%', fig.height=5}
df <- filter(s_md, grepl('TRB:1', chains), !is.na(az.annot.filtered)) %>%
  dplyr::rename(celltype = az.annot.filtered) %>%
  mutate(TRB = stringr::str_extract(string = clonotype, pattern = 'TRB:[A-Z]+')) %>%
  group_by(sample_id, celltype) %>%
  summarise(cells = n(), clonotypes = length(unique(TRB)), clonality = 1 - norm_entropy(TRB))

filter(df, cells > 9) %>%
  ggplot(aes(sample_id, clonality, size = cells)) + geom_point() +
  facet_wrap(~ celltype, ncol = 5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position="bottom") +
  xlab('Sample ID') + ylab('Clonality') +
  ggtitle('Clonality per cell type (10 or more cells) per sample (TCRB only)')
```

### Enrichment in clusters / cell types {.tabset}

#### Clusters

Check enrichment in clusters
```{r}
coi <- filter(s_md, !is.na(clonotype), has_TRAB) %>%
  group_by(clonotype) %>%
  summarise(n = n()) %>%
  filter(n > 1) %>%
  pull(clonotype)
enrich_res <- filter(s_md, !is.na(clonotype) & clonotype %in% coi) %>% 
  group_by(orig.ident, clonotype, integration_cluster, .drop = FALSE) %>%
  summarise(enrich_test(x_val = clonotype[1], 
                        y_val = integration_cluster[1], 
                        x_vec = s_md$clonotype[s_md$orig.ident == orig.ident[1]], 
                        y_vec = s_md$integration_cluster[s_md$orig.ident == orig.ident[1]])) %>%
  ungroup()
enrich_sum <- mutate(enrich_res, p_val_adjust = p.adjust(p_val)) %>%
  filter(p_val_adjust <= 0.05) %>%
  arrange(p_val_adjust)

DT::datatable(data = enrich_sum, rownames = FALSE) %>% 
  DT::formatRound(columns = c('odds_ratio'), digits = 3) %>%
  DT::formatSignif(columns = c('p_val', 'p_val_adjust'))
```

Heatmap of odds ratio for the top clonotypes. Shown are log2-odds-ratios limited to the range -5, 5.
```{r, fig.width=12, fig.height=7, out.width='100%'}
selected_ct <- filter(enrich_sum, rank(p_val_adjust) <= 30) %>%
  pull(clonotype) %>% unique()
odds_mat <- filter(enrich_res, clonotype %in% selected_ct) %>%
  mutate(log_odds_ratio = log2(scales::oob_squish_any(x = odds_ratio, range = c(1/32, 32))),
         clonotype = paste0(stringr::str_match(orig.ident, 'KB\\d+_'), clonotype)) %>%
  acast(formula = clonotype ~ integration_cluster, value.var = 'log_odds_ratio')
gplots::heatmap.2(odds_mat, scale = 'none', trace = 'none', margins = c(3,35), 
                  col = RColorBrewer::brewer.pal(n = 11, name = 'RdYlGn'))
```

In which clusters do we see enrichment most often?
```{r}
sort(table(enrich_sum$integration_cluster), decreasing = TRUE)
```

#### Azimuth cell types

Check Azimuth predicted cell types for clonotype enrichment
```{r}
enrich_res <- filter(s_md, !is.na(clonotype) & clonotype %in% coi) %>% 
  group_by(orig.ident, clonotype, az.annotation, .drop = FALSE) %>%
  summarise(enrich_test(x_val = clonotype[1], 
                        y_val = az.annotation[1], 
                        x_vec = s_md$clonotype[s_md$orig.ident == orig.ident[1]], 
                        y_vec = s_md$az.annotation[s_md$orig.ident == orig.ident[1]])) %>%
  ungroup()
enrich_sum <- mutate(enrich_res, p_val_adjust = p.adjust(p_val)) %>%
  filter(p_val_adjust <= 0.05) %>%
  arrange(p_val_adjust)

DT::datatable(data = enrich_sum, rownames = FALSE) %>% 
  DT::formatRound(columns = c('odds_ratio'), digits = 3) %>%
  DT::formatSignif(columns = c('p_val', 'p_val_adjust'))
```

Heatmap of odds ratio for the top clonotypes. Shown are log2-odds-ratios limited to the range -5, 5.
```{r, fig.width=12, fig.height=7, out.width='100%'}
selected_ct <- filter(enrich_sum, rank(p_val_adjust) <= 30) %>%
  pull(clonotype) %>% unique()
odds_mat <- filter(enrich_res, clonotype %in% selected_ct) %>%
  mutate(log_odds_ratio = log2(scales::oob_squish_any(x = odds_ratio, range = c(1/32, 32))),
         clonotype = paste0(stringr::str_match(orig.ident, 'KB\\d+_'), clonotype)) %>%
  acast(formula = clonotype ~ az.annotation, value.var = 'log_odds_ratio')
gplots::heatmap.2(odds_mat, scale = 'none', trace = 'none', margins = c(8,25), 
                  cexRow = 1, cexCol = 0.9,
                  col = RColorBrewer::brewer.pal(n = 11, name = 'RdYlGn'))
```

In which cell types do we see enrichment most often?
```{r}
sort(table(enrich_sum$az.annotation), decreasing = TRUE)
```



### In context of predicted cell types

Tables below show the predicted cell type and whether the cell is annotated with at least one A or B chain and no other chains. Values of NA mean no celltype or clonotype annotation is available.

```{r}
table(s_md$az.annotation, s_md$only_TRAB, useNA = 'always')
```

Show how the top clonotypes are distributed among the predicted cell types

```{r}
tmp <- filter(s_md, clonotype %in% top_clono) %>%
  group_by(az.annotation, clonotype) %>%
  summarise(n = n()) %>%
  acast(formula = az.annotation ~ clonotype, value.var = 'n')
print(t(tmp))
```


## TRD and TRG

```{r, fig.width=10, fig.height=4.5, out.width='100%'}
p1 <- filter(s_md, has_TRDG) %>%
  ggplot(aes(sample_id, fill = chains)) + geom_bar() + ylab('Cells') +
  ggtitle('Cells with at least one D or G chain') +
  scale_fill_discrete(name = 'Chain counts') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

show(p1)
```

Clonotype diversity based on cells with exactly one D and one G chains observed
```{r}
tmp <- filter(s_md, chains == 'TRD:1;TRG:1') %>%
  group_by(sample_id) %>%
  summarise(n = n(), sdi(clonotype))
p2 <- ggplot(tmp, aes(n, exp(sdi))) + 
  geom_abline(slope = 1, color = 'gray80') +
  geom_point() +
  geom_label_repel(aes(label = sample_id), max.overlaps = Inf) +
  xlab('Cells') + ylab('Effective number of clonotypes') 
show(p2)
```

Show  clonotypes that have been ranked in the top 5 and have been observed more than two times
```{r, fig.width=10, out.width='100%'}
top_clono <- filter(freq_dg, rank <= 5, n > 2) %>% pull(clonotype) %>% unique()
filter(freq_dg, clonotype %in% top_clono) %>% 
  DT::datatable(rownames = FALSE) %>% DT::formatRound(columns = 'freq', digits = 3)
```
```{r, out.width='100%', fig.width=10, fig.height=7}
p2 <- filter(freq_dg, clonotype %in% top_clono) %>% ungroup() %>%
  mutate(sample_id = factor(sample_id, levels = rev(unique(sample_id))),
         clonotype = factor(clonotype, levels = unique(clonotype))) %>%
  ggplot(aes(clonotype, sample_id, fill = freq)) +
  geom_raster() +
  scale_fill_continuous(name = 'Frequency') +
  theme(axis.text.x = element_text(angle = 66, hjust = 1)) +
  ggtitle('Clonotype frequency heatmap')
show(p2)
```

### In context of predicted cell types 

Tables below show the predicted cell type and whether the cell is annotated with at least one D or G chain and no other chains. Values of NA mean no celltype or clonotype annotation is available.

```{r}
table(s_md$az.annotation, s_md$only_TRDG, useNA = 'always')
```

Show how the top clonotypes are distributed among the predicted cell types

```{r}
tmp <- filter(s_md, clonotype %in% top_clono) %>%
  group_by(az.annotation, clonotype) %>%
  summarise(n = n()) %>%
  acast(formula = az.annotation ~ clonotype, value.var = 'n')
print(t(tmp))
```


## IGH, IGK and IGL

```{r, fig.width=10, fig.height=4.5, out.width='100%'}
p1 <- filter(s_md, has_IG) %>%
  ggplot(aes(sample_id, fill = chains)) + geom_bar() + ylab('Cells') +
  ggtitle('Cells with at least one H, K, or L chain') +
  scale_fill_discrete(name = 'Chain counts') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
show(p1)
```

Show only clonotypes that have been ranked in the top 5 and have been observed more than one time
```{r}
top_clono <- filter(freq_ig, rank <= 5, n > 1) %>% pull(clonotype) %>% unique()
filter(freq_ig, clonotype %in% top_clono) %>% 
  DT::datatable(rownames = FALSE) %>% DT::formatRound(columns = 'freq', digits = 3)
```
```{r, out.width='100%', fig.width=10, fig.height=7}
p2 <- filter(freq_ig, clonotype %in% top_clono) %>% ungroup() %>%
  mutate(sample_id = factor(sample_id, levels = rev(unique(sample_id))),
         clonotype = factor(clonotype, levels = unique(clonotype))) %>%
  ggplot(aes(clonotype, sample_id, fill = freq)) +
  geom_raster() +
  scale_fill_continuous(name = 'Frequency') +
  theme(axis.text.x = element_text(angle = 66, hjust = 1))
show(p2)
```

### In context of predicted cell types

Tables below show the predicted cell type and whether the cell is annotated with at least one H, K or L chain and no other chains. Values of NA mean no celltype or clonotype annotation is available.

```{r}
table(s_md$az.annotation, s_md$only_IG, useNA = 'always')
```

Show how the top clonotypes are distributed among the predicted cell types

```{r}
tmp <- filter(s_md, clonotype %in% top_clono) %>%
  group_by(az.annotation, clonotype) %>%
  summarise(n = n()) %>%
  acast(formula = az.annotation ~ clonotype, value.var = 'n')
print(t(tmp))
```


## iNKT and MAIT cells

> Cell Ranger labels cells as potential iNKT and MAIT cells based on a combination of 1) matching the appropriate V and J genes for the TCR alpha, 2) matching the appropriate V genes for the TCR beta, and 3) matching known V(D)J junctions (CDR3s) from the literature.

### iNKT evidence in context of predicted cell types 

Tables below show the predicted cell type and whether the cell has evidence of being iNKT

```{r}
table(s_md$az.annotation, s_md$inkt_evidence, useNA = 'always')
```

### MAIT evidence in context of predicted cell types 

Tables below show the predicted cell type and whether the cell has evidence of being MAIT

```{r}
table(s_md$az.annotation, s_md$mait_evidence, useNA = 'always')
```

## References

[10x immune profiling glossary](https://support.10xgenomics.com/single-cell-vdj/software/pipelines/latest/algorithms/glossary)

[10x immune profiling annotation algorithm](https://support.10xgenomics.com/single-cell-vdj/software/pipelines/latest/algorithms/annotation#cdr3)

[10x iNKT and MAIT annotation](https://support.10xgenomics.com/single-cell-vdj/software/pipelines/latest/algorithms/single-cell-vdj/software/pipelines/latest/algorithms/inkt-mait)

## Runtime and session info

Runtime: `r time_diff(start_time)`

Session Info
```{r}
sessionInfo()
```
